Variability in upwelling patterns across the Benguela upwelling systems

### 2019-10-09 Write_up/Paper1_writeup.Rmd
updated

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@Mead; @schlegel]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarred; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

@@

knitr::include_graphics("../Figures/OISST.png")
```

@@

knitr::include_graphics("../Figures/SACTN.png")
```

@@

knitr::include_graphics("../Figures/CMC.png")
```

@@

This study aimed to investigate patterns occurring within upwelling events at various different spatial intervals from the coastline, as well as to test whether the same patterns are discernible in the various datasets. As seawater temperature is known to have a large influence on species distribution and productivity (Bolton 2010; Smit et al. 2013), thus it is important to understand not only how temperatures vary at different distances from the coastline but what contributing factors to ecosystem maintenance the intensity of these events have across these distance intervals from the coastline. Of particular interest to this study; the Benguela upwelling system, which is distinct in that it is one of four most productive regions in the world.

@@

We would be amiss to point out some weaknesses inherent in the SACTN dataset. Temperature
measurement along the South African coastline started in the 1970s, and has been inconsistent in
terms of the instruments used, the modifications to styles and methods, calibration, as well as site locations themselves (Smit et al. 2013; Schlegel and Smit 2016). Additionally, older records within some datasets, such as the SACTN dataset, may have been lost or are unreliable since the metadata for these are no longer available. These metadata are essential as their absence prevents the understanding of the influence regarding the utilised instruments on temperature recordings. It can thus be speculated that some measurements may not accurately represent well-validated temperature data due to the instrumentation used; however, we mitigated these effects by selecting only the most reliable and longest duration data, which other current and on-going studies, such as that by Williamson et al. have suggested are most suitable for the problem at hand.

Data in the SACTN are compiled from multiple sources, in addition to being acquired by various means, thus creating considerable heterogeneity; however, efforts are
currently underway to homogenise these datasets (Williamson et al., in progress). The underwater
temperature recorders (UTRs) used to collect data in the SACTN dataset expressed a lower number of NA values as compared to the data collected with hand thermometers. As such, this may have influenced the overall time series dataset (Schlegel and Smit, 2016). The level of precision at which data were collected also influenced the length of the time series needed. Time series in which temperature is collected at a precision of 0.5°C may require another 24 months of recordings to precisely detect long term variation (Schlegel and Smit, 2016). The average length of the thermometer time series component of the SACTN dataset was 346 months whereas the average length of UTR time series was less than half of that. With the extent of these differences in length being so severe, even once correcting for potential negative effects on the measurement precision of the thermometer collected time series, it was clear that thermometer data were more useful than that of UTRs. These influences may be expected to affect trend estimates of change derived from the data, however it is not anticipated that they have affected estimates of the variability in time and between sites in the way that they were used in this current study.

Irrespective, satellite-acquired SST records are useful to modern marine scientists. These data are often used to model and predict a wide range of oceanic and biological processes within open ocean ecosystems and habitats; however, these have only recently been used to study temperature variations influencing benthic organisms (Pearce, Faskel, and Hyndes, 2006). SST data acquired by satellites deviate from coastal in situ collected seawater temperatures for a variety of reasons (Smit et al. 2013). The SST data presented the broader-scale situation along the coast, therefore utilised here as a forcing boundary hypothesised here as influencing coastal temperatures. Analysis shows that the offshore SSTs (various sources) at a distance between 30km and 50km often do not match the detected events present within the in situ dataset along the west coast of South African.

>>>>

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@Mead; @schlegel; @schlegel2016]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarred; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

@@

knitr::include_graphics("../Figures/OISST.pdf")
```

@@

knitr::include_graphics("../Figures/SACTN.pdf")
```

@@

knitr::include_graphics("../Figures/CMC.pdf")
```

@@

This study aimed to investigate patterns occurring within upwelling events at various different spatial intervals from the coastline, as well as to test whether the same patterns are discernible in the various datasets. As seawater temperature is known to have a large influence on species distribution and productivity [@Bolton2010; @smit2013], thus it is important to understand not only how temperatures vary at different distances from the coastline but what contributing factors to ecosystem maintenance the intensity of these events have across these distance intervals from the coastline. Of particular interest to this study; the Benguela upwelling system, which is distinct in that it is one of four most productive regions in the world.

@@

We would be amiss to point out some weaknesses inherent in the SACTN dataset. Temperature measurement along the South African coastline started in the 1970s, and has been inconsistent in terms of the instruments used, the modifications to styles and methods, calibration, as well as site locations themselves [@Smit2013; @schlegel2016]. Additionally, older records within some datasets, such as the SACTN dataset, may have been lost or are unreliable since the metadata for these are no longer available. These metadata are essential as their absence prevents the understanding of the influence regarding the utilised instruments on temperature recordings. It can thus be speculated that some measurements may not accurately represent well-validated temperature data due to the instrumentation used; however, we mitigated these effects by selecting only the most reliable and longest duration data, which other current and on-going studies, such as that by Williamson et al. have suggested are most suitable for the problem at hand.

Data in the SACTN are compiled from multiple sources, in addition to being acquired by various means, thus creating considerable heterogeneity; however, efforts are currently underway to homogenise these datasets (Williamson et al., in progress). The underwater temperature recorders (UTRs) used to collect data in the SACTN dataset expressed a lower number of NA values as compared to the data collected with hand thermometers. As such, this may have influenced the overall time series dataset [@schlegel2016]. The level of precision at which data were collected also influenced the length of the time series needed. Time series in which temperature is collected at a precision of 0.5°C may require another 24 months of recordings to precisely detect long term variation [@schlegel2016]. The average length of the thermometer time series component of the SACTN dataset was 346 months whereas the average length of UTR time series was less than half of that. With the extent of these differences in length being so severe, even once correcting for potential negative effects on the measurement precision of the thermometer collected time series, it was clear that thermometer data were more useful than that of UTRs. These influences may be expected to affect trend estimates of change derived from the data, however it is not anticipated that they have affected estimates of the variability in time and between sites in the way that they were used in this current study.

Irrespective, satellite-acquired SST records are useful to modern marine scientists. These data are often used to model and predict a wide range of oceanic and biological processes within open ocean ecosystems and habitats; however, these have only recently been used to study temperature variations influencing benthic organisms (Pearce, Faskel, and Hyndes, 2006). SST data acquired by satellites deviate from coastal in situ collected seawater temperatures for a variety of reasons [@smit2013]. The SST data presented the broader-scale situation along the coast, therefore utilised here as a forcing boundary hypothesised here as influencing coastal temperatures. Analysis shows that the offshore SSTs (various sources) at a distance between 30km and 50km often do not match the detected events present within the in situ dataset along the west coast of South African.
